Species turnover reveal hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-17 Manuscript.Rmd
Diskussion überarbeitet.

---
title: Recovery of mountain plant communities in response to reductions in Nitrogen emissions is hidden by other drivers of global change
preprint: false
author: 
  - name: Tobias Roth
    affiliation: 1, 2
    corresponding: true
    email: t.roth@unibas.ch
  - name: Lukas Kohli
    affiliation: 2
  - name: Beat Rihm
    affiliation: 3
  - name: Reto Meier
    affiliation: 4    
affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest,Fabrikstrasse 14, 3012 Bern, Switzerland 
  - code: 4
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition. We infered different drivers of community change (i.e. Nitrogen deposition, climate warming, land-use change) in Swiss mountain hay meadows. The data were obtained from the Swiss biodiversity monitoring.

@@

## Potential effects of reduction in Nitrogen emmissions {-}
```{r Ndepchange, fig.cap="(a) The nitrogen deposition in 2010 and (b) the change in Nitrogen deposition between 2000 and 2010 along the elevational gradient of the study sites."}
p1 <- ggplot(sites, aes(x = Elevation * 500 + 1000, y = NTOT_2000)) +
  geom_point(cex = 0.8) +
  geom_smooth(method = loess) +
  geom_abline(intercept = 0, slope = 0, lty = 3) +
  labs(title = "(a) Nitrogen deposition in 2010", 
       y = expression(paste("Nitrogen deposition (kg ", ha^{-1}," ", yr^{-1}, ")")), 
       x = "Elevation (m)") 
p2 <- ggplot(sites, aes(x = Elevation * 500 + 1000, y = (NTOT_2010 - NTOT_2000) )) +
  geom_point(cex = 0.8) +
  geom_smooth(method = loess)  +
  labs(title = "(b) Temporal change (2000 to 2010)", 
       y = expression(paste(Delta, "Nitrogen deposition (kg ", ha^{-1}," ", yr^{-1}, ")")), 
       x = "Elevation (m)") +
  geom_hline(yintercept = 0, lty = 2) 
multiplot(p1, p2, cols = 2)
```

Nitrogen deposition decreased with increasing elevation (Fig. \@ref(fig:Ndepchange)a). In 2010 only `r 100 * mean(sites$NTOT_2010 < 10)`% of sites had a Nitrogen deposition rate of less than 10 kg N ha$^{-1}$ yr$^{-1}$, which lied all above 1000m. Between 2000 and 2010 the Nitrogen deposition decreased in average around 2 kg N ha$^{-1}$ yr$^{-1}$ with slightly larger net decreases at lower elevation (Fig. \@ref(fig:Ndepchange)a).

@@

# Discussion {-}
Although Nitrogen deposition considerably declined between 2000 and 2015 (Fig. \@ref(fig:Ndepchange)), we could detect only weak shifts in plant community structure during the same time period (Table \@ref(tab:communitytrendstab). While the slight increase in average temperature value suggests that plant communities adopted to increasing temperatures, the constant aveage nutrients value suggests that the decrease in Nitrogen deposition did not yet affected plant communities. However, this apparent stability in community structure was accompanied by marked temporal turnover in species identities. It seems unlikely that this temporal turnover can entirly be explained by methodological issues such as overlooked species. First, pseudo-turnover of species entities in two independent surveys of the same site during the same season was smaller than the observed temporal turnover between two surveys from different years. Second, spatial variation of turnover showed patterns that can hardly be explained by methodological issues. For instance species turnover varied along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition. Taken together, our results adds to the increasing evidence that contemporary communities are often relatively stability at community level, but this apparent stability is often accompanied by marked turnover of species [@Vellend2013, @Dornelas2014, @Hillebrand2018].

Species communities are shaped by a range of factors, including deterministic processes such as environmental filtering or competitive interactions [@Goetzenberger2012, @Janevcek2013]. Such factors select for species with specific characteristics. Community assembly theories thus suggest that the factors that drive change in species communities could be inferred from comparing the charactereistics of disappearing or colonizing species with random expectation [@Chase2011]. We adopted this idea and compared the indicator values of species that disappeared or colonized a site with the indicator values from randomly chosen species from the same site. We found that the nutrient values of colonization species showed largest deviations from random expectation suggesting that Nitrogen deposition or other factors that change the nutrient content of soils were the most important drivers of the change in species composition over the last 15 years in Swiss mountain hay meadows.

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might overrule Nitrogen depositin effects on community structure. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizating species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizating species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high Nitrogen deposition. This suggests that it takes more time to replace eutrophic by oligotrophic species than replacing oligotrophic by eutrophic species. Taken toghether, these reasons might explain why the response of mountain hay medow communities responded stronger to to climate warming affects than to Nitrogen reduction, altough the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition. They assume that the spatial variation in species richness (or other metrics of community structure) arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of N-deposition [@Clark2008, @Storkey2015]. In an recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richnes of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load and revealed very similar results as the spatial approach. Our results are thus validating the space for time approach.

Hower, the results of Fig. \@ref(fig:) also shows that the results of spatial comparisons must be interpreted carefully. When we used the same model to infer how species richness is decreasing along the Nitrogen deposition gradient at differnt time points, except for the first two study periods the decrease became steeper (more negative) over time. Our first interpretation was that the Nitrogen deposition effect became stronger over time. This contraticted our prediction that the nitrogen deposition should become weaker over time since Nitrogen deposition decreased. However, we then realized that species turnover was highest in low deposition sites (Tab. \@ref(tab:turnovertab)) where colonizationg species were species with below average indicator values for Nutrients. It seems that the decrease in Nitrogen deposition resulted in higher numbers of oligotrophic species at sites with low deposition, thus leading to a steeper decline in oligotrophic species richness from sites with low to sites with high Nitrogen deposition.

# Conclusions {-}
Our study adds to the understanding of contemporary biodiversity change that is currently not not well understood (Magurran et al 2018.pdf). Our results support the notion of Hillebrand_et_al-2018.pdf that biodiversity monitoring programmes that quantify species turnover will generate a far more reliable understanding of the biotic response to changing environments than programmes solely tracking species richness.

>>>>

---
title: Species turnover reveal hidden effects of decreasing Nitrogen deposition in mountain hay meadows
preprint: false
author: 
  - name: Tobias Roth
    affiliation: 1, 2
    corresponding: true
    email: t.roth@unibas.ch
  - name: Lukas Kohli
    affiliation: 2
  - name: Beat Rihm
    affiliation: 3
  - name: Reto Meier
    affiliation: 4    
affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest,Fabrikstrasse 14, 3012 Bern, Switzerland 
  - code: 4
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity of many habitats. The recent introduction of cleaner technologies in Switzerland has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition. We infered different drivers of community change (i.e. Nitrogen deposition, climate warming, land-use change) in Swiss mountain hay meadows. The data were obtained from the Swiss biodiversity monitoring.

@@

## Potential effects of reduction in Nitrogen emmissions {-}
```{r Ndepchange, fig.cap="(a) The nitrogen deposition in 2000 and (b) the change in Nitrogen deposition between 2000 and 2010 along the elevational gradient of the study sites."}
p1 <- ggplot(sites, aes(x = Elevation * 500 + 1000, y = NTOT_2000)) +
  geom_point(cex = 0.8) +
  geom_smooth(method = loess) +
  geom_abline(intercept = 0, slope = 0, lty = 3) +
  labs(title = "(a) Nitrogen deposition in 2010", 
       y = expression(paste("Nitrogen deposition (kg ", ha^{-1}," ", yr^{-1}, ")")), 
       x = "Elevation (m)") 
p2 <- ggplot(sites, aes(x = Elevation * 500 + 1000, y = (NTOT_2015 - NTOT_2000) )) +
  geom_point(cex = 0.8) +
  geom_smooth(method = loess)  +
  labs(title = "(b) Temporal change (2000 to 2015)", 
       y = expression(paste(Delta, "Nitrogen deposition (kg ", ha^{-1}," ", yr^{-1}, ")")), 
       x = "Elevation (m)") +
  geom_hline(yintercept = 0, lty = 2) 
multiplot(p1, p2, cols = 2)
```

Nitrogen deposition decreased with increasing elevation (Fig. \@ref(fig:Ndepchange)a). In 2000 only `r 100 * mean(sites$NTOT_2000 < 10)`% of sites had a Nitrogen deposition rate of less than 10 kg N ha$^{-1}$ yr$^{-1}$, which lied all above 1000m. Between 2000 and 2015 the Nitrogen deposition decreased in average `r mean(sites$NTOT_2015 - sites$NTOT_2000)` $\pm$ `r sd(sites$NTOT_2015 - sites$NTOT_2000)` kg N ha$^{-1}$ yr$^{-1}$ with slightly higher net decreases at lower elevation (Fig. \@ref(fig:Ndepchange)a).

@@

# Discussion {-}
Although Nitrogen deposition considerably declined between 2000 and 2015 (Fig. \@ref(fig:Ndepchange)), we could detect only weak shifts in plant community structure during the same time period (Table \@ref(tab:communitytrendstab)). While the slight increase in average temperature value suggests that plant communities adopted to increasing temperatures, the constant aveage nutrients value suggests that the decrease in Nitrogen deposition did not yet affected plant communities. However, this apparent stability in community structure was accompanied by marked temporal turnover in species identities. It seems unlikely that this temporal turnover can entirly be explained by methodological issues such as overlooked species. First, pseudo-turnover of species entities in independent surveys of the same site during the same season was smaller than the observed temporal turnover between two surveys from different years. Second, spatial variation of turnover showed patterns that can hardly be explained by methodological issues. For instance species turnover varied along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition (Table \@ref(tab:turnovertab)). Taken together, our results adds to the increasing evidence that contemporary communities are often relatively stable at community level, but this apparent stability is often accompanied by marked turnover of species [@Vellend2013; @Dornelas2014; @Hillebrand2018].

Species communities are shaped by a range of factors, including deterministic processes such as environmental filtering or competitive interactions [@Goetzenberger2012; @Janevcek2013]. Such factors select for species with specific characteristics. Community assembly theories thus suggest that the factors that drive the composition of species in a community can be inferred from comparing the charactereistics of the species in the community with random expectation [@Chase2011]. We adopted this idea and compared the indicator values of species that disappeared or colonized a site with the indicator values from randomly chosen species from the same site. We found that the nutrient values of colonization species showed largest deviations from random expectation suggesting that Nitrogen deposition or other factors that change the nutrient content of soils were the most important drivers of the change in species composition over the last 15 years in Swiss mountain hay meadows.

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition in Swiss mountain hay meadows might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled  Nitrogen depositin effects on community structure. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizating species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizating species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high Nitrogen deposition. This suggests that it takes more time to replace eutrophic by oligotrophic species than replacing oligotrophic by eutrophic species. Taken toghether, these reasons might explain why the response of mountain hay medow communities responded stronger to to climate warming affects than to Nitrogen reduction, altough the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition [@Stevens2010gradient; chapter 4 in @deVries2015]. They assume that the spatial variation in species richness (or other metrics of community structure) arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of N-deposition in extant communities [@Clark2008; @Storkey2015]. In an recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richnes of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load. With this aproach we obtained very similar results as the spatial approach. Our results are thus validating the space for time approach.

Hower, the results of Fig. \@ref(fig:figconsequences) also shows that the results of spatial comparisons must be interpreted carefully. When we used the same model to infer how species richness is decreasing along the Nitrogen deposition gradient at differnt time points, except for the first two study periods the decrease became steeper (more negative) over time. Our first interpretation was that the Nitrogen deposition effect became stronger over time. This contraticted our prediction that the nitrogen deposition should become weaker over time since Nitrogen deposition decreased. However, we then realized that species turnover was highest in low deposition sites (Tab. \@ref(tab:turnovertab)) where the colonizating species were species with below average indicator values for Nutrients. It seems that the decrease in Nitrogen deposition resulted in oligotrophic species replacing eutrophic species especially at sites with low deposition, thus leading to a steeper decline in oligotrophic species richness from sites with low to sites with high Nitrogen deposition. The increasingly steeper decline over time in the eutrophic species richness along the spatial Nitrogen deposition gradient is thus a sign of the recovery of plant communities in low deposition sites rather than a sign that the Nitrogen deposition effects became stronger over time.

# Conclusions {-}
Comparing the indicator values of colonizating and desappearing species with random expecation we found that oligotrphic species are currently more likeley to colonize mountain hay meadows than oligotrophic species, which might be the results of the recently decreasing atmospeheric Nitrogen deposition. However, our results also caution that re-transfering mountain hay maeadows to oligotrophic species communities after high nitrogen depostion might take much longer than transfering a and oligotrophic into an eutrophic community. This is because eutrophic species have high local survival probabilities also when Nitrogen deposition becomes lower again. All these results were, however, hidden when only inferring the temporal trends of species communities. Our study thus adds to the understanding of contemporary biodiversity change that is currently not well understood [@Magurran2018], and it supports the notion of @Hillebrand2018 that inferring species turnover will generate a far more reliable understanding of the biotic response to changing environments than solely tracking community composition.
